---
title: "Covariate exploration and model selection"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 7
)
```

# Setup & data loading

A systematic covariate exploration was performed to identify metadata variables
driving variation in the RNA-seq data, collapse redundant covariates, and
select the best-fitting design for differential expression testing. The
analysis proceeded through normalisation (TMM + voom with two-pass
`duplicateCorrelation`), PCA to identify dominant axes of variation,
mixed-model testing of PC--metadata associations, hierarchical clustering
of correlated covariates, stepwise model building, and model comparison on
fit quality and DE concordance.

## Load pipeline functions and libraries

```{r load-functions}
source("../R/normalise_counts.R")
source("../R/test_pc_associations.R")
source("../R/plot_pca.R")
source("../R/cluster_covariates.R")
source("../R/suggest_models.R")
source("../R/compare_voom_models.R")
source("../R/select_model.R")

library(limma)
library(edgeR)
library(ggplot2)
library(patchwork)
library(tibble)
library(dplyr)
library(lmerTest)
library(lme4)
```

## Input data format

The `eval=FALSE` chunk below shows how real count and metadata files would be
loaded. For this analysis, simulated data were used instead (see next section).

```{r your-data, eval=FALSE}
counts   <- as.matrix(read.delim("path/to/counts.tsv", row.names = 1))
metadata <- read.delim("path/to/metadata.tsv")

# Ensure sample order matches
stopifnot(identical(colnames(counts), rownames(metadata)))
```

## Simulated demo data

A simulated dataset with known covariate structure was used to demonstrate the
pipeline end-to-end. The simulation produced 12 subjects x 2 conditions
(Ctrl/Treat) x 2 timepoints (T1/T2) = 48 samples, with covariates wired to
specific PCs:

| Covariate            | Expected signal                          |
|----------------------|------------------------------------------|
| `group`/`condition`  | Strong effect on a top PC                |
| `rna_quality`        | Moderate--strong effect on a top PC      |
| `batch` / `sex`      | Confounded; shared effect on a top PC    |
| `age`                | Weaker continuous effect                 |
| `library_size_proxy` | Noise (negative control, not significant)|

```{r simulate-data}
set.seed(42)

n_genes    <- 10000
n_subjects <- 12
conditions <- c("Ctrl", "Treat")
timepoints <- c("T1", "T2")

# -- Metadata --
metadata <- expand.grid(
  subject   = paste0("S", seq_len(n_subjects)),
  condition = conditions,
  timepoint = timepoints,
  stringsAsFactors = FALSE
)
metadata$sample_id <- paste(metadata$subject, metadata$condition,
                            metadata$timepoint, sep = "_")
rownames(metadata) <- metadata$sample_id

# Group label (condition x timepoint)
metadata$group <- factor(paste0(metadata$condition, ".", metadata$timepoint),
                         levels = c("Ctrl.T1", "Ctrl.T2",
                                    "Treat.T1", "Treat.T2"))

# Subject-level covariates (constant within subject)
subj_info <- data.frame(
  subject = paste0("S", seq_len(n_subjects)),
  sex     = c("F", "F", "M", "F", "M", "M",
              "F", "M", "M", "F", "M", "F"),
  batch   = c("A", "A", "B", "A", "B", "B",
              "A", "B", "B", "A", "B", "A"),
  age     = c(25, 32, 45, 28, 51, 39, 27, 44, 50, 30, 48, 35),
  stringsAsFactors = FALSE
)
metadata <- merge(metadata, subj_info, by = "subject", sort = FALSE)
rownames(metadata) <- metadata$sample_id

# Sample-level covariates
metadata$rna_quality       <- round(rnorm(nrow(metadata), mean = 8, sd = 1.5), 1)
metadata$library_size_proxy <- round(runif(nrow(metadata), 20, 40), 1)

# Reorder rows to match sample_id column
metadata <- metadata[order(metadata$sample_id), ]

# -- Counts matrix --
# Build log-expression matrix, then convert to counts.
# This gives precise control over covariate effects on PCs.
n_samples  <- nrow(metadata)
base_mean  <- rnorm(n_genes, mean = 6, sd = 1.5)        # log2 scale
log_expr   <- matrix(base_mean, nrow = n_genes, ncol = n_samples)

# Subject random intercept (intra-subject correlation)
subject_effects <- matrix(
  rnorm(n_genes * n_subjects, sd = 0.3),
  nrow = n_genes, ncol = n_subjects,
  dimnames = list(NULL, paste0("S", seq_len(n_subjects)))
)
for (i in seq_len(n_samples)) {
  log_expr[, i] <- log_expr[, i] + subject_effects[, metadata$subject[i]]
}

# group effect -> PC1: large fold changes in 2000 genes for Treat samples
group_genes <- 1:2000
treat_idx   <- metadata$condition == "Treat"
group_fc    <- rnorm(length(group_genes), mean = 0, sd = 1.5)
log_expr[group_genes, treat_idx] <- log_expr[group_genes, treat_idx] +
  outer(group_fc, rep(1, sum(treat_idx)))

# rna_quality -> PC2: proportional shift in 2000 genes
rna_genes <- 2001:4000
rna_z     <- scale(metadata$rna_quality)[, 1]
rna_coef  <- rnorm(length(rna_genes), mean = 0, sd = 0.8)
log_expr[rna_genes, ] <- log_expr[rna_genes, ] +
  outer(rna_coef, rna_z)

# batch/sex (confounded) -> PC3: shift in 1500 genes
batch_genes <- 4001:5500
batch_z     <- ifelse(metadata$batch == "B", 1, -1)
batch_coef  <- rnorm(length(batch_genes), mean = 0, sd = 0.7)
log_expr[batch_genes, ] <- log_expr[batch_genes, ] +
  outer(batch_coef, batch_z)

# age -> PC4: weaker continuous effect on 800 genes
age_genes <- 5501:6300
age_z     <- scale(metadata$age)[, 1]
age_coef  <- rnorm(length(age_genes), mean = 0, sd = 0.4)
log_expr[age_genes, ] <- log_expr[age_genes, ] +
  outer(age_coef, age_z)

# library_size_proxy: NO effect (negative control)

# Add residual noise
log_expr <- log_expr + matrix(rnorm(n_genes * n_samples, sd = 0.5),
                               nrow = n_genes, ncol = n_samples)

# Convert to counts via NB sampling around the log-expression means
mu <- 2^log_expr
mu[mu < 1] <- 1
counts <- matrix(
  rnbinom(n_genes * n_samples, mu = as.vector(mu), size = 20),
  nrow = n_genes, ncol = n_samples,
  dimnames = list(paste0("gene_", seq_len(n_genes)), metadata$sample_id)
)
counts[counts < 0] <- 0L
storage.mode(counts) <- "integer"

cat("Counts matrix:", nrow(counts), "genes x", ncol(counts), "samples\n")
cat("Metadata columns:", paste(names(metadata), collapse = ", "), "\n")
```


# Normalisation

Lowly-expressed genes were filtered with `edgeR::filterByExpr`, TMM
normalisation was applied, and both `voom` and `voomWithQualityWeights` were run with two-pass
`duplicateCorrelation` to account for the repeated-measures blocking structure.

```{r normalise}
norm <- normalise_counts(counts, group = metadata$group, block = metadata$subject)
```

```{r normalise-summary}
cat("Genes retained after filtering:", sum(norm$kept_genes), "/",
    length(norm$kept_genes), "\n")
cat("Block correlation (voom):               ",
    round(norm$block_correlation, 4), "\n")
cat("Block correlation (voomWithQualityWeights):",
    round(norm$block_correlation_sw, 4), "\n")
cat("\nSample weight summary:\n")
print(summary(norm$sample_weights))
```


# PCA

PCA was run on the voom-normalised expression matrix to identify the major axes
of variation across samples. Genes were centred and scaled prior to
decomposition with `prcomp`.

```{r pca}
pca <- prcomp(t(norm$voom), center = TRUE, scale. = TRUE)
```

```{r scree-plot, fig.width=6, fig.height=4}
var_explained <- summary(pca)$importance[2, 1:10] * 100
scree_df <- data.frame(
  PC       = factor(paste0("PC", 1:10), levels = paste0("PC", 1:10)),
  Variance = var_explained
)

ggplot(scree_df, aes(PC, Variance)) +
  geom_col(fill = "#4575B4") +
  labs(y = "% Variance Explained", title = "Scree Plot") +
  theme_minimal()
```


# PCA score plots (PC1 vs PC2)

Associations between the first 10 PCs and all metadata variables were tested
using mixed models. The default PC1 vs PC2 plane was then visualised coloured by
each variable. Stars in panel titles reflect the strength of association.

```{r test-associations}
pc_assoc <- test_pc_associations(
  pca, metadata[, !names(metadata) %in% c("sample_id")],
  n_pcs = 10,
  block = "subject",
  sample_weights = norm$sample_weights,
  plot = FALSE
)
```

```{r pca-score-plots-default, fig.width=12, fig.height=10}
pca_plots <- plot_pca(
  pca,
  metadata[, !names(metadata) %in% c("sample_id")],
  pc_assoc
)
pca_plots$combined_plot
```


# PC--metadata association heatmap

The heatmap displayed $-\log_{10}(p_\text{adj})$ for each variable--PC pair.
Grey tiles were non-significant; coloured tiles indicated progressively
stronger associations, with stars marking significance levels (\* *p* < 0.05,
\*\* *p* < 0.01, \*\*\* *p* < 0.001). Because the experiment used a
repeated-measures design, all associations were tested with linear mixed models
including a `(1|subject)` random intercept.

```{r heatmap, fig.width=10, fig.height=6}
pc_assoc$plot
```

```{r association-table}
pc_assoc$results |>
  dplyr::filter(p_adj < 0.05) |>
  dplyr::arrange(p_adj)
```

**Findings:**

- Variables associated with multiple PCs indicated broad technical confounding
  (e.g. `rna_quality`).
- Variables on the dominant PCs (PC1, PC2) reflected biological signal
  (e.g. `group`) or major batch effects.
- Sample-weight associations informed whether `voomWithQualityWeights` was
  warranted.


# Covariate clustering

Correlated covariates can cause multicollinearity when included together in a
model. Pairwise associations were computed (Spearman for continuous--continuous,
$\sqrt{\eta^2}$ for numeric--categorical, Cramer's V for
categorical--categorical), covariates were clustered hierarchically, and one
representative per cluster was selected based on PC-association score. `group` was excluded
as the variable of interest; `subject` was excluded as the blocking variable.

```{r cluster-covariates, fig.width=10, fig.height=7}
clust <- cluster_covariates(
  metadata[, !names(metadata) %in% c("sample_id")],
  pc_assoc = pc_assoc,
  pca      = pca,
  block    = "subject",
  exclude  = c("group")
)
clust$plot
```

```{r cluster-table}
clust$clusters
```

**Findings:**

- The dendrogram grouped covariates by similarity; the dashed line marked the
  clustering threshold.
- `*` next to a variable name denotes the cluster representative.
- `batch` and `sex` clustered together as expected given their confounding.
  Only the representative was carried forward for model building.


# PCA score plots (per-variable optimal PCs)

The default PC1 x PC2 view missed variables whose effects lay on lower PCs.
Each variable was re-plotted on its two most-associated PCs to give a clearer
picture of where each covariate acted.

```{r pca-per-variable, fig.width=12, fig.height=10}
pca_per_var <- plot_pca(
  pca,
  metadata[, !names(metadata) %in% c("sample_id")],
  pc_assoc,
  per_variable_pcs = TRUE
)
pca_per_var$combined_plot
```

**Findings:**

- Compared with the default PC1 x PC2 plots above, variables like
  `rna_quality` now appeared on PC2 x PC3 or similar, revealing structure
  invisible in the first two components.


# Suggest candidate models

Significant covariates were ranked by a variance-weighted score
($\sum R^2 \times \text{PC variance proportion}$ across significant PCs) and
candidate design matrices were built stepwise, checking for rank deficiency at each
addition. The filtered PC associations from the clustering step were used so
that redundant covariates were excluded.

```{r suggest-models}
suggested <- suggest_models(
  pc_assoc     = clust$pc_assoc_filtered,
  base_formula = ~ 0 + group,
  metadata     = metadata,
  block        = "subject",
  pca          = pca
)
```

```{r covariate-ranking}
suggested$covariate_ranking
```

```{r candidate-models}
for (m in suggested$models) {
  cat(m$name, ":", deparse(m$formula), "\n")
}
```

```{r sample-weights-flag}
if (suggested$sample_weights_flag) {
  cat("Sample weights are significantly associated with PC variation.\n",
      "voomWithQualityWeights will be compared in the next step.\n")
} else {
  cat("Sample weights are not significantly associated with any PC.\n")
}
```


# Compare voom models

For each candidate design, both `voom` and `voomWithQualityWeights` were fit
with two-pass `duplicateCorrelation`. Models were then compared on median RMSE,
median R^2^, DE gene counts, and concordance metrics (logFC correlation, rank
correlation, Jaccard index). The four-panel diagnostic plot showed RMSE and
R^2^ densities, DE counts, and a logFC scatter.

```{r compare-models, results="asis", fig.width=10, fig.height=8}
contrast    <- "groupTreat.T1 - groupCtrl.T1"
comparisons <- list()

for (m in suggested$models) {
  cat("\n\n##", m$name, "\n\n")

  comparisons[[m$name]] <- compare_voom_models(
    dge      = norm$dge,
    design   = m$design,
    contrast = contrast,
    block    = metadata$subject
  )

  cat("\n**Summary:**\n\n")
  print(knitr::kable(comparisons[[m$name]]$summary, digits = 4))
  cat("\n")
  print(comparisons[[m$name]]$plot)
  cat("\n")
}
```

**Findings:**

- Lower median RMSE indicated better fit; higher median R^2^ indicated more
  variance explained by the design.
- The logFC scatter fell tightly along the diagonal where models agreed,
  confirming high concordance.
- Models with worse fit but more DE genes were flagged as potentially
  inflating false positives.


# Select best model

`select_model()` picked the best voom strategy (voom vs
voomWithQualityWeights) within each design based on RMSE and R^2^, then
compared across designs. A 1% parsimony threshold favoured simpler models
unless a more complex model offered a meaningful improvement.

```{r select-model}
selection <- select_model(comparisons)
```

```{r selection-summary}
selection$design_summary
```

```{r recommendation}
message("Recommended design:  ", selection$design)
message("Recommended strategy: ", selection$recommended)
message("Reason: ", selection$reason)
```


# Selection diagnostics

A multi-panel diagnostic figure was produced to aid review of the model
selection process. Panels included an UpSet plot of DE gene overlap across
designs, a pairs plot of genome-wide logFC consistency, fit quality metrics,
per-gene RMSE distributions, p-value calibration histograms, and a summary
of the selection path.

```{r selection-diagnostics, fig.width=14, fig.height=18}
diag <- plot_selection_diagnostics(comparisons, selection)
diag$plot
```


# Summary & next steps

The pipeline recommended **`r selection$design`** with
**`r selection$recommended`** as the best-fitting model for this dataset.
The fitted model objects were extracted for downstream differential expression
analysis:

```{r next-steps, eval=FALSE}
# Extract the fitted model objects
efit      <- selection$model$efit
top_table <- selection$model$top_table

# Apply significance thresholds
de_results <- limma::decideTests(efit, p.value = 0.05, lfc = 0)
summary(de_results)

# Downstream: volcano plots, gene set enrichment, pathway analysis, etc.
```
